Detection and replication of epistasis influencing transcription in humans

### 2014-01-07 docs/manuscript/manuscript.tex
references updated, manuscript has now been accepted. woo

\documentclass{article}
\usepackage[superscript,biblabel]{cite}
\usepackage{graphicx}
\usepackage{float}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage{authblk}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{lscape}
% \usepackage{lineno}
\usepackage{hyperref}
% \linenumbers

@@

\graphicspath{{../../analysis/images/}}
\bibliographystyle{naturemag}

@@

Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics. Though often demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been reported in other species \cite{Carlborg2006}, few examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical and computational issues \cite{Cordell2009}. Here we show that, using advanced computation \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with 7339 gene expression levels measured in peripheral blood, we found 501 significant pairwise interactions between common SNPs influencing the expression of 238 genes ($p < 2.91 \times 10^{-16}$). Replication of these interactions in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$. Forty-four of the genetic interactions are located within 2Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of segregating common polymorphisms interacting to influence human traits.

@@

% \subsection{Introduction}
In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently and cumulatively to the mean of a trait. This framework has been successful in identifying thousands of associations \cite{Visscher2012}. But to date, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}, there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}. Beyond the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example, increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. Thus, with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants. Because our focus was to ascertain the extent to which instances of epistasis arises from natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes and like most complex diseases, these expression traits are typically heritable \cite{Schadt2003}. But unlike complex diseases, genetic associations with gene expression commonly have very large effect sizes that explain large proportions of the genetic variance\cite{Powell2013}, making them good candidates to search for epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions. First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood ($1.03 \times 10^{15}$ statistical tests, family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods). Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $0.05 \leq \alpha \leq 0.14$, Methods, Supplementary Figure \ref{fig:type1}). Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). We used strict quality control measures to avoid statistical associations being driven by technical artifacts (Methods). However it remains possible that unexplained technical artifacts may have led to the significant discovery interactions. Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/501$, Table \ref{tab:bonferroni}). To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A1 \times A2$, $A1 \times D2$, $D1 \times A2$, $D1 \times D2$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods). Sign concordance between the discovery and both replication datasets was observed in 22 out of the 30 significantly replicated interactions (expected value $= 7.5$ under the null hypothesis of no interactions, $p = 3.76 \times 10^{-8}$).

In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions where $p$-values are assumed to be uniformly distributed ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}). It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistatic effects in independent samples is difficult in practice due to LD (Methods).

% \subsection{Patterns of epistasis}
Though seldom the focus of association studies, SNPs with known main effects are often tested for $A \times A$ genetic interactions \cite{Cordell2009}, but our analysis suggests this is unlikely to be the best strategy for its detection. The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \cite{Powell2013} (439 out of 501, Methods). Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects. Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst $A \times A$, $A \times D$, $D \times A$ and $D \times D$ at the discovery stage ($p = 0.22$ for departure from expectation). This is not surprising because these patterns of epistasis used for statistical decomposition are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \cite{Cockerham1954}.

Of the discovery interactions, 26 were \emph{cis}-\emph{cis} acting (within 1Mb of the transcription start site, mean distance between SNPs was 0.53Mb), 462 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting. We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP. For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an effect (Supplementary Figure \ref{fig:MBNL1_3D}). Each of these interactions has evidence for replication in at least one dataset and six are significantly replicated at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}). We see similar epistatic networks involving multiple (eight or more) \emph{trans}-acting SNPs for other gene expresson levels too, for example TMEM149 (Supplementary Figure \ref{fig:TMEM149_3D}), NAPRT1 (Supplementary Figure \ref{fig:NAPRT1_3D}), TRAPPC5 (Supplementary Figure \ref{fig:TRAPPC5_3D}), and CAST (Supplementary Figure \ref{fig:CAST_3D}). We observed that from pedigree analysis these five gene expression phenotypes had non-additive variance component estimates within the 95th percentile of the 17,994 gene expression phenotypes that were analysed previously \cite{Powell2013} (Supplementary Table \ref{tab:vc}, Methods).

% \subsection{Functional annotation}
In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment (Methods). We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,tri-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}). There was significant enrichment for \emph{cis}-acting SNPs in haematopoietic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing). However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues). This difference between \emph{cis} and \emph{trans} SNPs suggests different roles in epistatic interactions where tissue specificity is provided by the \emph{cis} SNPs. There is also enrichment for \emph{cis}-SNPs to be localised in regions with regulatory genomic features as measured by chromatin states \cite{Hoffman2012} (Supplementary Figure \ref{fig:snplocations}).

We also demonstrate physical organisation of interacting loci within the cell, suggesting a mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}. We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 5Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Rieder2012}.

% \subsection{Contribution of epistatic variance}
Quantifying the importance of epistasis in complex traits in humans remains an open question. Here we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance. Taking results from our previously published eQTL \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the estimated additive variance of a locus was $2.1\%$ or greater. Thus, we can infer that the number of instances of large additive effects is significantly greater than the number of instances of large epistatic effects.

In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Taking all additive effects detected in Powell \emph{et al} (2012) that have additive variance explaining $2.1\%$ or greater of phenotypic variance, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$. By contrast, the estimated epistatic variance from the interacting SNPs detected in this study on average explain a total of $0.22\%$ of phenotypic variance, approximately ten times lower than the estimated additive variance. There are several caveats to this comparison which we discuss in the Methods.

@@

\section*{Methods Summary}
We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs. Each individual had gene expression levels measured in peripheral blood at 7,339 probes representing 6,158 RefSeq genes (significant expression in $\geq 90\%$ of individuals). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. We used permutation analysis to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects. All SNP pairs were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair. The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood Fehrmann \cite{Fehrmann2011}, $n=1240$; EGCUT \cite{Metspalu2004}, ($n=891$). A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.

@@

G.H., J.E.P., P.M.V., and G.W.M. conceived and designed the study. G.H., J.E.P., K.S., H-J.W., and J.Y. performed the analysis. T.E. and A.M. provided the EGCUT data. A.K.H., A.F.M., G.W.M., N.G.M., and J.E.P. provided the BSGS data. H-J.W. and L.F. provided the Fehrmann data. G.H. and J.E.P. wrote the manuscript with the participation of all authors.

@@

\bibliography{refs}
\clearpage

>>>>

\documentclass{article}
\usepackage{amsmath}
% \usepackage[superscript,biblabel]{cite}
\usepackage{graphicx}
\usepackage{float}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage{authblk}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{lscape}
% \usepackage{lineno}
\usepackage{verbatim}
\usepackage{hyperref}
% \linenumbers
% \usepackage{bibunits}
% \usepackage{splitbib}
\usepackage[style=nature, sorting=none]{biblatex}
\addbibresource{refs.bib}

\DeclareBibliographyCategory{Category1}
\addtocategory{Category1}{Hill2008a}
\addtocategory{Category1}{Costanzo2010}
\addtocategory{Category1}{Carlborg2006}
\addtocategory{Category1}{Strange2010}
\addtocategory{Category1}{Cordell2009}
\addtocategory{Category1}{Hemani2011}
\addtocategory{Category1}{Lieberman-Aiden2009}
\addtocategory{Category1}{Visscher2012}
\addtocategory{Category1}{Carlborg2004}
\addtocategory{Category1}{Weinreich2006}
\addtocategory{Category1}{Breen2012}
\addtocategory{Category1}{Weir2008}
\addtocategory{Category1}{Marchini2005}
\addtocategory{Category1}{LangoAllen2010}
\addtocategory{Category1}{Schadt2003}
\addtocategory{Category1}{Powell2013}
\addtocategory{Category1}{Powell2012}
\addtocategory{Category1}{Fehrmann2011}
\addtocategory{Category1}{Metspalu2004}
\addtocategory{Category1}{Preininger2013}
\addtocategory{Category1}{Cockerham1954}
\addtocategory{Category1}{Ho2004}
\addtocategory{Category1}{Trynka2013}
\addtocategory{Category1}{Hoffman2012}
\addtocategory{Category1}{Lan2012}
\addtocategory{Category1}{Rieder2012}
\addtocategory{Category1}{Crow2010}
\addtocategory{Category1}{Hemani2013}
\addtocategory{Category1}{Evans2011}
\addtocategory{Category1}{Bloom2013}

\DeclareBibliographyCategory{Category2}
\addtocategory{Category2}{Medland2009}
\addtocategory{Category2}{Aulchenko2007}
\addtocategory{Category2}{Yang2011}
\addtocategory{Category2}{Westra2011}
\addtocategory{Category2}{Williams1976}
\addtocategory{Category2}{Alvarez-Castro2008}
\addtocategory{Category2}{Koch2007}
\addtocategory{Category2}{Rietveld2013}
\addtocategory{Category2}{Stormo2000}
\addtocategory{Category2}{HoSui2005}

@@

\graphicspath{{../../analysis/images/}}
% \bibliographystyle{naturemag}

@@

Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \supercite{Carlborg2004} and contributes to their variation \supercite{Hill2008a,Crow2010} is a fundamental question in evolution and human genetics. Though often demonstrated in artificial gene manipulation studies in model organisms \supercite{Costanzo2010,Bloom2013}, and some examples have been reported in other species \supercite{Carlborg2006}, few examples exist for epistasis amongst natural polymorphisms in human traits \supercite{Strange2010,Evans2011}. Its absence from empirical findings may simply be due to low incidence in the genetic control of complex traits \supercite{Hill2008a,Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical and computational issues \supercite{Cordell2009}. Here we show that, using advanced computation \supercite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with 7339 gene expression levels measured in peripheral blood, we found 501 significant pairwise interactions between common SNPs influencing the expression of 238 genes ($p < 2.91 \times 10^{-16}$). Replication of these interactions in two independent data sets \supercite{Metspalu2004,Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$. Forty-four of the genetic interactions are located within 2Mb of regions of known physical chromosome interactions \supercite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of segregating common polymorphisms interacting to influence human traits.

@@

% \subsection{Introduction}
In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently and cumulatively to the mean of a trait. This framework has been successful in identifying thousands of associations \supercite{Visscher2012}. But to date, though its contribution to phenotypic variance is frequently the subject of debate \supercite{Carlborg2004,Hill2008a,Crow2010}, there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \supercite{Strange2010, Evans2011}. Beyond the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \supercite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \supercite{Weinreich2006} and speciation \supercite{Breen2012}.

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example, increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \supercite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \supercite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \supercite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. Thus, with small genetic effect sizes, as is expected in most complex traits of interest \supercite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \supercite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants. Because our focus was to ascertain the extent to which instances of epistasis arises from natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes and like most complex diseases, these expression traits are typically heritable \supercite{Schadt2003}. But unlike complex diseases, genetic associations with gene expression commonly have very large effect sizes that explain large proportions of the genetic variance\supercite{Powell2013}, making them good candidates to search for epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \supercite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions. First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood ($1.03 \times 10^{15}$ statistical tests, family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods). Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $0.05 \leq \alpha \leq 0.14$, Methods, Supplementary Figure \ref{fig:type1}). Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). We used strict quality control measures to avoid statistical associations being driven by technical artifacts (Methods). However it remains possible that unexplained technical artifacts may have led to the significant discovery interactions. Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \supercite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \supercite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/501$, Table \ref{tab:bonferroni}). To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A1 \times A2$, $A1 \times D2$, $D1 \times A2$, $D1 \times D2$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods). Sign concordance between the discovery and both replication datasets was observed in 22 out of the 30 significantly replicated interactions (expected value $= 7.5$ under the null hypothesis of no interactions, $p = 3.76 \times 10^{-8}$).

In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions where $p$-values are assumed to be uniformly distributed ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}). Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \supercite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}). It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistatic effects in independent samples is difficult in practice due to LD (Methods).

% \subsection{Patterns of epistasis}
Though seldom the focus of association studies, SNPs with known main effects are often tested for $A \times A$ genetic interactions \supercite{Cordell2009}, but our analysis suggests this is unlikely to be the best strategy for its detection. The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \supercite{Powell2013} (439 out of 501, Methods). Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects. Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst $A \times A$, $A \times D$, $D \times A$ and $D \times D$ at the discovery stage ($p = 0.22$ for departure from expectation). This is not surprising because these patterns of epistasis used for statistical decomposition are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \supercite{Cockerham1954}.

Of the discovery interactions, 26 were \emph{cis}-\emph{cis} acting (within 1Mb of the transcription start site, mean distance between SNPs was 0.53Mb), 462 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting. We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP. For example, MBNL1 (involved in RNA modification and regulation of splicing \supercite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an effect (Supplementary Figure \ref{fig:MBNL1_3D}). Each of these interactions has evidence for replication in at least one dataset and six are significantly replicated at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}). We see similar epistatic networks involving multiple (eight or more) \emph{trans}-acting SNPs for other gene expresson levels too, for example TMEM149 (Supplementary Figure \ref{fig:TMEM149_3D}), NAPRT1 (Supplementary Figure \ref{fig:NAPRT1_3D}), TRAPPC5 (Supplementary Figure \ref{fig:TRAPPC5_3D}), and CAST (Supplementary Figure \ref{fig:CAST_3D}). We observed that from pedigree analysis these five gene expression phenotypes had non-additive variance component estimates within the 95th percentile of the 17,994 gene expression phenotypes that were analysed previously \supercite{Powell2013} (Supplementary Table \ref{tab:vc}, Methods).

% \subsection{Functional annotation}
In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment (Methods). We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,tri-methylation (H3K4me3) chromatin marks, in 34 cell types \supercite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}). There was significant enrichment for \emph{cis}-acting SNPs in haematopoietic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing). However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues). This difference between \emph{cis} and \emph{trans} SNPs suggests different roles in epistatic interactions where tissue specificity is provided by the \emph{cis} SNPs. There is also enrichment for \emph{cis}-SNPs to be localised in regions with regulatory genomic features as measured by chromatin states \supercite{Hoffman2012} (Supplementary Figure \ref{fig:snplocations}).

We also demonstrate physical organisation of interacting loci within the cell, suggesting a mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \supercite{Lieberman-Aiden2009}. We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \supercite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 5Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \supercite{Rieder2012}.

% \subsection{Contribution of epistatic variance}
Quantifying the importance of epistasis in complex traits in humans remains an open question. Here we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance. Taking results from our previously published eQTL \supercite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the estimated additive variance of a locus was $2.1\%$ or greater. Thus, we can infer that the number of instances of large additive effects is significantly greater than the number of instances of large epistatic effects.

In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \supercite{Hill2008a}. Taking all additive effects detected in Powell \emph{et al} (2012) that have additive variance explaining $2.1\%$ or greater of phenotypic variance, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$. By contrast, the estimated epistatic variance from the interacting SNPs detected in this study on average explain a total of $0.22\%$ of phenotypic variance, approximately ten times lower than the estimated additive variance. There are several caveats to this comparison which we discuss in the Methods.

@@

\section*{Methods Summary}
We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \supercite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs. Each individual had gene expression levels measured in peripheral blood at 7,339 probes representing 6,158 RefSeq genes (significant expression in $\geq 90\%$ of individuals). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. We used permutation analysis to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects. All SNP pairs were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair. The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood Fehrmann \supercite{Fehrmann2011}, $n=1240$; EGCUT \supercite{Metspalu2004}, ($n=891$). A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.

@@

G.H., J.E.P., P.M.V., and G.W.M. conceived and designed the study. G.H., J.E.P., K.S., H-J.W., and J.Y. performed the analysis. T.E. and A.M. provided the EGCUT data. A.K.H., A.F.M., G.W.M., N.G.M., and J.E.P. provided the BSGS data. G.G. provided the CHDWB data. H-J.W. and L.F. provided the Fehrmann data. G.H. and J.E.P. wrote the manuscript with the participation of all authors.

@@

% \bibliography{refs}{References 1}
\printbibliography[category=Category1]
\clearpage

\section*{Online methods}

\section{Discovery data}

\subsection{Data description}

The Brisbane Systems Genetics Study (BSGS) comprises 846 individuals of European descent from 274 independent families \supercite{Powell2012}. DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland. Full details of genotyping procedures are given in Medland et al. \supercite{Medland2009} Standard quality control (QC) filters were applied and the remaining 528,509 autosomal SNPs were carried forward for further analysis.

Gene expression profiles were generated from peripheral blood collected with PAXgene TM tubes (QIAGEN, Valencia, CA) using Illumina HT12-v4.0 bead arrays. The Illumina HT-12 v4.0 chip contains 47,323 probes, although some probes are not assigned to RefSeq genes. We removed any probes that did not match the following criteria: contained a SNP within the probe sequence with MAF $ > 0.05$ within 1000 genomes data; did not map to a listed RefSeq gene; were not significantly expressed (based on a detection $p$-value $< 0.05$) in at least 90\% of samples. After this stringent QC 7339 probes remained for 2D-eQTL mapping. These data are accessible through GEO Series accession number GSE53195.

\subsection{Normalisation}

Gene expression profiles were normalised and adjusted for batch and polygenic effects. Profiles were first adjusted for raw background expression in each sample. Expression levels were then adjusted using quantile and $\log_2$ transformation to standardise distributions between samples. Batch and polygenic effects were adjusted using the linear model

\begin{equation}
y = \mu + \beta_{1}c + \beta_{2}p + \beta_{3}s + \beta_{4}a + g + e
\label{eq:lm}
\end{equation}
where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively, fitted as fixed effects; and $g$ is a random additive polygenic effect with a variance covariance matrix 
\begin{equation}
G_{jk} = \left \{ 
\begin{array}{ll}
\sigma _a ^2             &          j = k \\ 
2\phi _{jk} \sigma _a ^2& 			j \neq k \\
\end{array} \right.
\end{equation}
The parameter $\sigma_a^2$ is the variance component for additive background genetic. Here, we are using family based pedigree information rather than SNP based IBD to account for relationships between individuals and so $\phi _{jk}$ is the kinship coefficient between individuals $j$ and $k$. The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero. Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan to remove any skewness and avoid results being driven by outliers. The GenABEL package for R was used to perform the normalisation \supercite{Aulchenko2007}.

\section{Exhaustive 2D-eQTL analysis}

\subsection{Two stage search}

We used epiGPU \supercite{Hemani2011} software to perform an exhaustive scan for pairwise interactions, such that each SNP is tested against all other SNPs for statistical association with the expression values for each of the 7339 probes. This uses the massively parallel computational architecture of graphical processing units (GPUs) to speed up the exhaustive search. For each SNP pair there are 9 possible genotype classes. We treat each genotype class as a fixed effect and fit an 8 \emph{d.f.} $F$-test to test the following hypotheses:
